---
title: "Brief investigation on sampling from exponential distribution "
author: "Ilya Semenov"
date: "March 29, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Overview

This is a brief illustration of [Central Limit Theorem](https://en.wikipedia.org/wiki/Exponential_distribution) in action. Means and variances of 1000 simulated samples of length 40 exponential averages were compared with analytically obtained formulas for exponential mean and variance. Used exponential distribution with Lambda parameter 0.2.

##Simulations

According to theory mean of exponential distribution with LAMBDA == 0.2 equals
```{r echo=TRUE}
lambda <- 0.2
1/lambda
```

So according to Central Limit Theorem (CLT) we can expect that mean of 1000 40-sized samples will approximately be equal to 5. Let check it.

```{r echo=TRUE}
mns <- NULL
set.seed(1001)
samplesize = 40
for (i in 1:1000) {
    mns <- c(mns, mean(rexp(samplesize, lambda)))
}
mean(mns)
```

It proves our theoretical estimate is correct.

##Sample mean vs theoretical mean

According to CLT sample means must be approximately normally distributed around theoretical mean of exponential distribution with lambda parameter 0.2, that equals 5.
On a plot below blue vertical line represents theoretical mean.

```{r echo=TRUE}
hist(mns, main = 'Histogram of sample means frequencies', 
     xlab = 'Sample means values', col = 'lightgreen')
abline(v=5, lwd=5, col='blue')
```

##Sample variance vs theoretical variance

Theoretical standard deviation for exponential distribution samples of size 40 equals approximately 0.79. 
```{r}
thsd <- 1/lambda/sqrt(samplesize)
thsd
```

Theoretical variance equals approximately 0.625.
```{r}
thsd^2
```

Let check if simulated samples standard deviation and variance matche with those values.

```{r}
sd(mns)
var(mns)
```

So expected and obtained standard deviation and variance match.

##Is distribution close to normal?

To answer this question we can graphically compare obtained samples mean distribution with normal distribution with parameters mean = 5, sd = 0.79

```{r echo=TRUE}
hist(mns, breaks = 20, probability = TRUE, ylim = c(0, .6), 
     main = 'Histogram of sample means density', 
     xlab = 'Sample means values', 
     col = 'lightgreen')
curve(dnorm(x, mean=5, sd = .79), lwd = 3, add = TRUE, col = 'blue')
```

So as seen from the figure sample means distribution fits with normal distribution perfectly.